---
title: "h3sdm workflow for a single model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{h3sdm workflow for a single model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates a basic workflow for species distribution modeling 
using the `h3sdm` package.

## Load packages
```{r}
library(h3sdm)
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
```

## 1. Area of interest
```{r}
cr <- cr_outline_c
plot(st_geometry(cr))
```

## 2. Species occurrence data

The package includes a precomputed presence/pseudo-absence dataset for 
*Silverstoneia flotator* at H3 resolution 7.
```{r}
data(records)
head(records)
table(records$presence)
```
```{r}
ggplot() +
  theme_minimal() +
  geom_sf(data = records, aes(fill = presence)) +
  scale_fill_manual(
    name = "Presence/Absence",
    values = c("red", "blue"),
    labels = c("Absence", "Presence")
  )
```

## 3. Environmental predictors
```{r}
bio <- terra::rast(system.file("extdata", "bioclim_current.tif", package = "h3sdm"))
h7 <- h3sdm_get_grid(cr, res = 7)
bio_predictors <- h3sdm_extract_num(bio, h7)
predictors <- h3sdm_predictors(bio_predictors) |>
  dplyr::select(h3_address, bio1, bio12, bio15, geometry)
```

## 4. Further information

For a complete workflow including model fitting, spatial cross-validation, 
predictions, and feature importance, see the 
[full tutorial](https://manuelspinola.github.io/h3sdm/articles/h3sdm_single_model.html).
