---
title: "h3sdm workflow for multiple models"
author: "Manuel Spínola"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates a complete SDM workflow for a single species using
multiple model types with `h3sdm`. We cover data preparation with a hexagonal
grid, environmentally stratified pseudo-absence generation, model fitting with
spatial cross-validation, performance comparison, predictions, and variable
importance analysis.

## Load packages

```{r}
library(h3sdm)
library(tidyverse)
library(tidymodels)
library(spatialsample)
library(sf)
library(terra)
library(DALEX)
library(DALEXtra)
library(ingredients)
library(workflowsets)
library(themis)
library(ranger)
library(xgboost)
library(stacks)
```

## 1. Define the Area of Interest

We start by defining the geographical area for modeling. Here we use Costa Rica
as an example. The file is included in the `h3sdm` package.

```{r}
cr <- cr_outline_c
```

## 2. Load Environmental Predictors

We use WorldClim historic bioclimatic variables for Costa Rica as environmental
predictors. The data is included in the `h3sdm` package.

```{r}
bio <- terra::rast(system.file("extdata", "bioclim_current.tif", package = "h3sdm"))
names(bio) <- gsub(".*bio_", "bio", names(bio))
```

## 3. Create the Hexagonal Grid

The hexagonal grid is the backbone of the `h3sdm` workflow. All subsequent
steps — predictor extraction, presence assignment, and pseudo-absence generation —
are built on top of it. At resolution 7, each hexagon covers approximately
514 ha, which is appropriate for many vertebrate species.

```{r}
h7 <- h3sdm_get_grid(cr, res = 7)
```

```{r}
ggplot() +
  geom_sf(data = h7)
```

## 4. Prepare Predictors

Environmental variables are extracted for every hexagon in the grid. Here we
use three WorldClim bioclimatic variables: Bio1 (Annual Mean Temperature),
Bio12 (Annual Precipitation), and Bio15 (Precipitation Seasonality).

```{r}
bio_predictors <- h3sdm_extract_num(bio, h7)
```

```{r}
predictors <- h3sdm_predictors(bio_predictors) |>
  dplyr::select(h3_address, bio1, bio12, bio15, geometry)
```

We can visualize one of the predictors, for example Bio1.

```{r}
ggplot() +
  theme_minimal() +
  geom_sf(data = predictors, aes(fill = bio1)) +
  scale_fill_viridis_c(option = "B")
```

## 5. Species Occurrence Data

Presence hexagons are generated using `h3sdm_pres()`, which downloads occurrence
records and assigns them to hexagons. Multiple records within the same hexagon
are consolidated into a single presence, reducing spatial sampling bias. Since
each hexagon represents an area (~514 ha at resolution 7), this better reflects
how organisms actually occupy space.

Pseudo-absences are then generated with `h3sdm_pa()`. They are placed outside
the known distribution — excluding hexagons with presences and their immediate
neighbors (`buffer_k = 1`) — and selected using k-means clustering in
environmental space. This ensures pseudo-absences represent the full range of
environmental conditions available in the study area. Since there are
approximately 100 presence hexagons at resolution 7, we request 300
pseudo-absences (3×).

```{r}
pres <- h3sdm_pres("Silverstoneia flotator", cr, res = 7, limit = 10000)
```

```{r}
records <- h3sdm_pa(pres, predictors, n_pseudoabs = 300)
```

```{r}
head(records)
```

```{r}
table(records$presence)
```

```{r}
ggplot() +
  theme_minimal() +
  geom_sf(data = records, aes(fill = presence))
```

## 6. Combine Records and Predictors

Merge species occurrence records with environmental predictors.

```{r}
dat <- h3sdm_data(records, predictors)
```

## 7. Spatial Cross-Validation

Define spatial blocks for cross-validation to account for spatial
autocorrelation.

```{r}
scv <- h3sdm_spatial_cv(dat, v = 5, repeats = 1)
```

```{r}
autoplot(scv)
```

## 8. Define Recipe and Models

Create a modeling recipe and specify multiple classification models. We start
with presence–absence data aggregated in hexagonal cells. The dataset is
imbalanced — approximately 100 presence hexagons and 300 pseudo-absence hexagons —
which can bias models toward predicting absences. We use
`step_downsample(presence)` from the `themis` package to correct for this.
After down-sampling, the dataset is balanced, improving model training and
evaluation while retaining all presence information.

```{r}
receta <- h3sdm_recipe(dat) |>
  themis::step_downsample(presence) |>
  step_dummy(all_nominal_predictors())
```

Now we define multiple models using the `parsnip` package from the tidymodels
framework.

```{r}
mod_log <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

mod_rf <- rand_forest() %>%
  set_engine("ranger") %>%
  set_mode("classification")

mod_xgb <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("classification")

my_models <- list(
  reg_log       = mod_log,
  random_forest = mod_rf,
  xgboost       = mod_xgb
)
```

## 9. Create Workflows

```{r}
wfs <- h3sdm_workflows(my_models, receta)
```

```{r}
wfs
```

## 10. Fit the Models

Before fitting, we extract the presence data from the dataset. This ensures
that metrics, cross-validation, and evaluation focus correctly on the locations
where the species is actually present.

```{r}
presence_data <- dat %>%
  dplyr::filter(presence == 1)
```

We fit the models using the spatial cross-validation scheme. Spatial CV
accounts for spatial autocorrelation by partitioning the data into spatially
distinct folds, providing a more realistic assessment of model performance
compared to random CV.

```{r}
several <- h3sdm_fit_models(wfs, scv, presence_data)
```

## 11. Evaluate and Compare Models

```{r}
compare <- h3sdm_compare_models(several)
compare
```

Three metrics are reported:

- **ROC AUC** (`roc_auc`) — evaluates the model's ability to discriminate
  between presence and pseudo-absence regardless of threshold. It is the
  standard metric for probabilistic classification.
- **Maximum TSS** (`tss_max`) — combines sensitivity and specificity into a
  single threshold-dependent value, showing how well the model predicts
  presences and absences simultaneously.
- **Boyce index** (`boyce`) — measures the model's ability to predict species
  distribution continuously and assesses whether areas with higher predicted
  values coincide with observed presences.

```{r}
ggplot(compare, aes(.metric, mean)) +
  geom_col(width = 0.03) +
  geom_point(size = 2) +
  facet_wrap(~model)
```

## 12. Select the Best Model and Make Predictions

```{r}
p_rf <- h3sdm_predict(several$models$random_forest, predictors)
```

```{r}
p_rf
```

## 13. Map

```{r}
ggplot() +
  theme_minimal() +
  geom_sf(data = p_rf, aes(fill = prediction)) +
  scale_fill_viridis_c(name = "Habitat \nsuitability", option = "B", direction = -1)
```

The map represents habitat suitability across hexagons. Values between 0 and 1
indicate the probability of the species being present in each hexagon.
Higher values indicate more suitable habitat; lower values indicate less
suitable habitat.

## 14. Model Interpretation: Feature Importance and Partial Dependence

We interpret the model to understand which predictors are most influential and
how they affect predictions. First, we extract the fitted random forest model.

```{r}
rf_fitted <- several$models$random_forest$final_model
```

Then we create an explainer object using the DALEX package.

```{r}
e <- h3sdm_explain(rf_fitted, data = dat)
```

### Feature Importance

We evaluate the importance of each predictor variable using permutation
importance. This method assesses how much model performance decreases when
the values of a predictor are randomly shuffled.

```{r}
predictors_to_evaluate <- setdiff(names(e$data), c("h3_address", "x", "y", "presence"))
```

```{r}
var_imp <- model_parts(
  explainer = e,
  variables = predictors_to_evaluate
)
```

```{r}
plot(var_imp)
```

### Partial Dependence Plots

Partial dependence plots (PDPs) show how the predicted outcome changes as a
single predictor varies, while averaging out the effects of all other predictors.

```{r}
pdp_rf <- partial_dependence(e, variables = c("bio12", "bio1", "bio15"))
```

```{r}
plot(pdp_rf)
```

## Conclusions

This vignette demonstrated a complete SDM workflow for multiple models using
`h3sdm`. The key steps included defining the study area, creating the hexagonal
grid, preparing environmental predictors, generating presence hexagons and
environmentally stratified pseudo-absences, fitting multiple models with spatial
cross-validation, comparing performance, making predictions, and assessing
variable importance. This workflow can be adapted to any species and study area
by adjusting the resolution, the number of pseudo-absences, and the set of
environmental predictors.
